Finiteness and Cardinality in Homotopy Type Theory

### 2020-01-10 paper.tex
done ish

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
\subsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

Figure~\ref{finite-classification} organises the predicates according to
their strength; i.e. how much information they provide about a conforming
type.
For instance, a proof that some type \(A\) is manifestly Bishop finite (the
strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
us that \(A\) is discrete (has decidable equality), and gives us a linear order
on the type.
A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
features: indeed, we will see examples of Kuratowski finite types which are not
even sets, never mind discrete ones.

\input{figures/finite-classification}

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.
Our main proof in this regard is that we can derive manifest Bishop finiteness
from Kuratowski finiteness plus a total order.

@@

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
    \begin{equation}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
    \begin{equation}
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
    \end{equation}
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
\end{rm-definition}
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \iff \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}

\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma} \label{split-enum-bool-top-bot}
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  Each of these types clearly has a finite number of elements (2, 1, and 0,
  respectively), and furthermore has a straightforward enumeration.
\end{proof}
\begin{rm-theorem} \label{split-enum-sigma}
  Split-enumerability is closed under \(\Sigma\).
  \begin{equation} \label{split-enum-sigma-clos-sig}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-sup}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction} \cite[definition
  3.11.1]{hottbook}; i.e. its membership proof exists, and all such proofs are
  equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}
By the definition of a contraction, we can always recover the underlying
membership proof, meaning that we can always derive split enumerability from
manifest Bishop finiteness.
\subsection{Equivalence}
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite[definition 4.4.1]{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function (called \(\text{uniques}\)) which filters
  out duplicates from lists of \(A\)s.
  This gives us our support list.
  To generate the cover proof it suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
\end{proof}
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means that we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}

We will use the circle as an example non-set type which is nonetheless
manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

@@

  We then need to convert a value of type \(\mathbf{Dec}(x \in \mathit{xs})\) to
  \(x \in \mathit{xs}\).
  We use the following to do that:
  \begin{equation}
    \text{recompute} : \mathbf{Dec}(A) \rightarrow \lVert A \rVert \rightarrow A
  \end{equation}
  The function works via case-analysis on \(\mathbf{Dec}\): in the true case, we
  return the proof; in the false case, we can apply the proof of negation under
  the truncation, and then extract \(\lVert \bot \rVert\) to \(\bot\), as
  \(\bot\) is a mere proposition, thus proving the goal via explosion.
\end{proof}
\subsection{Closure}
Since we don't have an equivalence with \(\mathbf{Fin}\), we don't get closure
under \(\Pi\).
\begin{rm-lemma}
  Manifest enumerability is closed under \(\Sigma\).
\end{rm-lemma}
\begin{proof}
  This closure proof is almost the same as theorem~\ref{split-enum-sigma}.
  The manipulation of the support lists can be carried over as-is; but the type
  of the cover proof has changed, so it will need to be updated.
  As it happens, the translation is straightforward: we effectively write the
  ``monadic'' version of the old function.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
  \begin{equation}
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}%
\end{rm-definition}%
\subsection{Closure}
The closure properties of cardinal finiteness are effectively the non-dependent
versions of manifest Bishop finiteness.
To see why, consider equation~\ref{split-enum-sigma-clos-sig}.
We can ``lift'' the proof (as a binary function) under a propositional
truncation, giving us equation~\ref{sigma-proof-lifted}, but that doesn't give us
the desired closure proof (equation~\ref{sigma-proof-desired}).
\begin{multicols}{2}
  \noindent
  \begin{equation} \label{sigma-proof-lifted}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \lVert \Pi (x : A) , \mathcal{B}(U(x)) \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
  \end{equation}
  \begin{equation} \label{sigma-proof-desired}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \Pi (x : A) , \lVert \mathcal{B}(U(x))  \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
  \end{equation}
\end{multicols}%
To remedy the mismatch we would need a function of the type:
\begin{equation}
  \left( \Pi(x : A) , \lVert \mathcal{B}(U(x)) \rVert \right) \rightarrow
  \lVert \Pi(x : A) , \mathcal{B}(U(x)) \rVert
\end{equation}
Unfortunately, this equation in particular is a form of the axiom of choice
\cite[equation 3.8.3]{hottbook}.
This leaves us with closure under only the non-dependent operations.
\begin{rm-lemma}
  Cardinal finiteness is closed under \(\times\), \(+\), and \(\rightarrow\).
\end{rm-lemma}
\begin{proof}
  All of these closure proofs can be lifted directly from their corresponding
  proofs on manifest Bishop finiteness.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{rm-theorem}
\begin{proof}
  We will use eliminator~\ref{elim-prop-prop} from definition~\ref{prop-trunc}.
  Manifest Bishop finiteness implies decidable equality already, so our task
  here is to prove that decidable equality itself is a proposition.

  We know that if a type \(A\) is a proposition, then the decision over that
  type is also a proposition.
  Then, via Hedberg's theorem, we know that any type with decidable equality is
  a set, meaning that paths in that type are themselves propositions.
  Therefore we can derive that a decision of equality on elements with decidable
  equality is a proposition, and by function extensionality we see that
  decidable equality is itself a proposition.
\end{proof}
\subsection{Cardinality}
\begin{rm-theorem}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{rm-theorem}
\begin{proof}
  Our task here is to ``pull out'' the cardinality of the set from under the
  propositional truncation.
  In effect, we need the following function:
  \begin{equation}
    \left\lVert \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right) \right\rVert \rightarrow
    \Sigma {(n : \mathbb{N})} , \left\lVert \mathbf{Fin}(n) \simeq A \right\rVert
  \end{equation}

  We will use eliminator~\ref{elim-prop-coh} from definition~\ref{prop-trunc}.
  We eliminate with the following function:
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{trunc-alg}

  To show that \(\text{alg}\) is coherently constant, we first notice that the
  second element of the output pair is propositionally truncated, meaning that
  it is trivially equal to any other element of the same type.
  Our task, then, simplifies to demonstrating that the first element of the
  output pair is coherently constant.
  \begin{equation}
    \Pi {(x : \Sigma {(n : \mathbb{N})} , \textbf{Fin}(n) \simeq A)} ,
    \Pi {(y : \Sigma {(m : \mathbb{N})} , \textbf{Fin}(m) \simeq A)} ,
    n \equiv m
  \end{equation}

  Notice that \(\mathbf{Fin}(n)\) and \(\mathbf{Fin}(m)\) are both equivalent to
  \(A\): we can join these proofs together, giving us the following:
  \begin{equation}
    \textbf{Fin}(n) \equiv \textbf{Fin}(m)
  \end{equation}
  All that remains now is to prove that \(\mathbf{Fin}\) is injective.
  Though the proof is surprisingly complex, it is a well-known puzzle in
  Martin-Löf type theory.
  Our proof does not differ significantly from standard approaches, so we will
  not detail it here.
\end{proof}
\subsection{Relation to Manifest Bishop Finiteness}
Cardinal finiteness tells us that there is an isomorphism between a type and
\(\mathbf{Fin}\); it just doesn't tell us \emph{which} isomorphism.
To take a simple example, \(\mathbf{Bool}\) has 2 possible isomorphisms with the
set \(\mathbf{Fin}(2)\): one where false maps to 0, and true to 1; and another
where false maps to 1 and true to 0.

@@

  Again, this proof is quite complex, so we only give a sketch here.

@@

\begin{definition}
\end{definition}

@@

  As with lists, it is cleaner, theory-wise, to define streams as a container.

@@

  Conceptually, a stream is like a list without an end.

@@

\subsubsection{\(\Sigma\) Closure}

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.5\linewidth}
    \begin{tikzcd}[sep=tiny,font=\footnotesize]
      (1,e) \ar[rdddd, out=-45, in=135] & (2,e) \ar[rdddd, out=-45, in=135] & (3,e) \ar[rdddd, out=-45, in=135] & (4,e) \ar[rdddd, out=-45, in=135] & (5,e)        \\
      (1,d) \ar[u]     & (2,d) \ar[u]    & (3,d) \ar[u]    & (4,d) \ar[u]    & (5,d) \ar[u] \\
      (1,c) \ar[u]     & (2,c) \ar[u]    & (3,c) \ar[u]    & (4,c) \ar[u]    & (5,c) \ar[u] \\
      (1,b) \ar[u]     & (2,b) \ar[u]    & (3,b) \ar[u]    & (4,b) \ar[u]    & (5,b) \ar[u] \\
      (1,a) \ar[u]     & (2,a) \ar[u]    & (3,a) \ar[u]    & (4,a) \ar[u]    & (5,a) \ar[u]
    \end{tikzcd}
    \caption{Cartesian}
    \label{cartesian}
  \end{subfigure}%
  \begin{subfigure}[b]{.5\linewidth}
    \begin{tikzcd}[sep=tiny,font=\footnotesize]
      (1,e) \ar[dr] & (2,e) \ar[dr]  & (3,e) \ar[dr]    & (4,e) \ar[dr] & (5,e) \\
      (1,d) \ar[dr] & (2,d) \ar[dr]  & (3,d) \ar[dr]    & (4,d) \ar[dr] & (5,d) \ar[u] \\
      (1,c) \ar[dr] & (2,c) \ar[dr]  & (3,c) \ar[dr]    & (4,c) \ar[dr] & (5,c) \ar[uul] \\
      (1,b) \ar[dr] & (2,b) \ar[dr]  & (3,b) \ar[dr]    & (4,b) \ar[dr] & (5,b) \ar[uuull, out=130, in=-50] \\
      (1,a) \ar[u]  & (2,a) \ar[uul, out=130, in=-50] & (3,a) \ar[uuull, out=130, in=-50] & (4,a) \ar[uuuulll, out=130, in=-50] & (5,a) \ar[uuuulll, out=130, in=-50]
    \end{tikzcd}
    \caption{Cantor}
    \label{cantor}
  \end{subfigure}

>>>>

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
\subsection{Strong Finiteness Predicates}
\input{figures/finite-classification}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

Figure~\ref{finite-classification} organises the predicates according to
their strength.
We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.
Our main proof in this regard is that we can derive manifest Bishop finiteness
from Kuratowski finiteness plus a total order.

@@

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  This definition is what is referred to as \verb+Listable+ in
\end{rm-definition}
\cite{firsovDependentlyTypedProgramming2015}.
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
    \begin{equation}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
    \begin{equation}
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
    \end{equation}
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
\end{rm-definition}
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \iff \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  After sufficient inlining, it emerges that our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma} \label{split-enum-bool-top-bot}
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  Each of these types clearly has a finite number of elements (2, 1, and 0,
  respectively), and furthermore has a straightforward enumeration.
\end{proof}
\begin{rm-theorem} \label{split-enum-sigma}
  Split-enumerability is closed under \(\Sigma\).
  \begin{equation} \label{split-enum-sigma-clos-sig}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-sup}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction} \cite[definition
  3.11.1]{hottbook}; i.e. its membership proof exists, and all such proofs are
  equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}
By the definition of a contraction, we can always recover the underlying
membership proof, meaning that we can always derive split enumerability from
manifest Bishop finiteness.
\subsection{Equivalence}
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite[definition 4.4.1]{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function (called \(\text{uniques}\)) which filters
  out duplicates from lists of \(A\)s.
  This gives us our support list.
  To generate the cover proof it suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
\end{proof}
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We will use three eliminators from \(\lVert A \rVert\) in this paper.
  \begin{enumerate}
  \item \label{elim-prop-prop} For any function \(A \rightarrow B\), where
    \(\text{isProp}(B)\), we have a function \(\lVert A \rVert \rightarrow B\).
  \item \label{elim-prop-monad} A special case of \ref{elim-prop-prop} implies
    that \(\lVert \cdot \rVert\) forms a monad: this means that we get the usual
    Monadic operators on \(\lVert \cdot \rVert\) (bind, pure, fmap, etc.).
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\) (\(\Pi {(x , y : A)} , f(x) \equiv f(y) \)).
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}

We will use the circle as an example non-set type which is nonetheless
manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\subsection{Relation to Split Enumerability}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

@@

  We then need to convert a value of type \(\mathbf{Dec}(x \in \mathit{xs})\) to
  \(x \in \mathit{xs}\).
  We use the following to do that:
    \text{recompute} : \mathbf{Dec}(A) \rightarrow \lVert A \rVert \rightarrow A
  The function works via case-analysis on \(\mathbf{Dec}\): in the true case, we
  return the proof; in the false case, we can apply the proof of negation under
  the truncation, and then extract \(\lVert \bot \rVert\) to \(\bot\), as
  \(\bot\) is a mere proposition, thus proving the goal via explosion.
\subsection{Closure}
Since we don't have an equivalence with \(\mathbf{Fin}\), we don't get closure
under \(\Pi\).
  Manifest enumerability is closed under \(\Sigma\).
  This closure proof is almost the same as theorem~\ref{split-enum-sigma}.
  The manipulation of the support lists can be carried over as-is; but the type
  of the cover proof has changed, so it will need to be updated.
  As it happens, the translation is straightforward: we effectively write the
  ``monadic'' version of the old function.
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}%
\end{rm-definition}%
\subsection{Closure}
The closure properties of cardinal finiteness are effectively the non-dependent
versions of manifest Bishop finiteness.
To see why, consider equation~\ref{split-enum-sigma-clos-sig}.
We can ``lift'' the proof (as a binary function) under a propositional
truncation, giving us equation~\ref{sigma-proof-lifted}, but that doesn't give us
the desired closure proof (equation~\ref{sigma-proof-desired}).
\begin{multicols}{2}
  \noindent
  \begin{equation} \label{sigma-proof-lifted}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \lVert \Pi (x : A) , \mathcal{B}(U(x)) \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
  \begin{equation} \label{sigma-proof-desired}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \Pi (x : A) , \lVert \mathcal{B}(U(x))  \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
\end{multicols}%
To remedy the mismatch we would need a function of the type:
  \left( \Pi(x : A) , \lVert \mathcal{B}(U(x)) \rVert \right) \rightarrow
  \lVert \Pi(x : A) , \mathcal{B}(U(x)) \rVert
Unfortunately, this equation in particular is a form of the axiom of choice
\cite[equation 3.8.3]{hottbook}.
This leaves us with closure under only the non-dependent operations.
  Cardinal finiteness is closed under \(\times\), \(+\), and \(\rightarrow\).
  All of these closure proofs can be lifted directly from their corresponding
  proofs on manifest Bishop finiteness.
\subsection{Strength}
The following two theorems are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014}, in much the same way as we have
done here.
Our contribution for this section is simply the formalisation.
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
  See \cite[Proposition 2.4.10]{yorgeyCombinatorialSpeciesLabelled2014}.
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \cite[Proposition 2.4.9]{yorgeyCombinatorialSpeciesLabelled2014}.
\subsection{Relation to Manifest Bishop Finiteness}
Cardinal finiteness tells us that there is an isomorphism between a type and
\(\mathbf{Fin}\); it just doesn't tell us \emph{which} isomorphism.
To take a simple example, \(\mathbf{Bool}\) has 2 possible isomorphisms with the
set \(\mathbf{Fin}(2)\): one where false maps to 0, and true to 1; and another
where false maps to 1 and true to 0.

@@

@@

\todo[inline]{Topos!}

@@

@@

@@

\subsubsection{Closure}

@@

\input{figures/pairing-functions.tex}
